---
title: "Group10_A2"
authors: "Jacqueline Gunawan, Byron Boukouvalas, Jacob Francis,
Taylor Smart"
format: 
  html: default
  pdf: default
editor: visual
execute:
  echo: FALSE
  warning: FALSE
  messages: FALSE
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  eval    = TRUE,
  fig.height = 4,
  fig.width  = 7
)

library(tidyverse)
library(kableExtra)
library(gridExtra)

theme_set(theme_minimal(base_size = 12))
trees <- read_csv("data/trees.csv")
```

## Task 1

### Summary Table:

```{r}
# Summary of age, ULE, and origin counts by precinct
summary_table <- trees |>
  group_by(precinct) |>
  summarise(
    n            = n(),
    age_mean     = round(mean(age), 1),
    age_median   = round(median(age), 1),
    age_sd       = round(sd(age), 1),
    age_range    = paste0(min(age), "–", max(age)),
    ule_mean     = round(mean(ule), 1),
    ule_median   = round(median(ule), 1),
    ule_sd       = round(sd(ule), 1),
    ule_range    = paste0(min(ule), "–", max(ule)),
    n_native     = sum(origin == "Native"),
    n_introduced = sum(origin == "Introduced")
  )

# Render table with grouped column headers
summary_table |>
  kable(col.names = c("Precinct", "N",
                      "Mean", "Median", "SD", "Range",
                      "Mean", "Median", "SD", "Range",
                      "Native", "Introduced"),
        caption = "Table 1: Summary of tree age, ULE, and species origin by precinct") |>
  kable_styling(full_width = TRUE) |>
  add_header_above(c(" " = 2,
                     "Age (years)" = 4,
                     "ULE (years)" = 4,
                     "Origin (count)" = 2))
```

### Distribution:

#### Overall:

```{r}
# Overall distribution of age
p_age <- ggplot(trees, aes(x = age, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "steelblue", alpha = 0.4) +
  labs(title = "Tree Age", x = "Age (years)", y = "Density")

# Overall distribution of ULE
p_ule <- ggplot(trees, aes(x = ule, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "steelblue", alpha = 0.4) +
  labs(title = "Useful Life Expectancy", x = "ULE (years)", y = "Density")

# Overall distribution of species origin
p_origin <- ggplot(trees, aes(x = origin, fill = origin)) +
  geom_bar(alpha = 0.7) +
  labs(title = "Species Origin", x = "Origin", y = "Count") +
  theme(legend.position = "none")

# Combine into one figure
grid.arrange(p_age, p_ule, p_origin, ncol = 3)
```

#### By precinct:

```{r}
# Age by precinct
p_age_pre <- ggplot(trees, aes(x = precinct, y = age, fill = precinct)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  labs(title = "Tree Age by Precinct", x = "Precinct", y = "Age (years)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# ULE by precinct
p_ule_pre <- ggplot(trees, aes(x = precinct, y = ule, fill = precinct)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  labs(title = "ULE by Precinct", x = "Precinct", y = "ULE (years)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Origin by precinct
p_origin_pre <- ggplot(trees, aes(x = precinct, fill = origin)) +
  geom_bar(position = "fill", alpha = 0.7) +
  labs(title = "Species Origin by Precinct", x = "Precinct",
       y = "Proportion", fill = "Origin") +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine
grid.arrange(p_age_pre, p_ule_pre, p_origin_pre, ncol = 2)
```

### Summary:

Across five distinct precincts, the urban forest comprises 630 trees with an average age of 30 years. While individual ages span from 1 to 146 years, the population leans heavily toward younger specimens. Southbank stands out as the primary home for the forest's "elderly" trees, which—while historically significant—generally possess a lower remaining Useful Life Expectancy (ULE) than their younger counterparts. Overall, the forest maintains a consistent ULE of approximately 50 years across all areas, and ULE remains broadly consistent across precincts, suggesting no single area faces a disproportionate replacement burden in the near term.

The botanical makeup is almost perfectly divided between native (313) and introduced (317) species, though the distribution is far from uniform. High-density areas like the CBD and Docklands favor introduced trees, whereas Carlton and East Melbourne feature a more integrated blend of local and foreign flora.

## Task 2
